---
title: "Ghost Flights -- pilot"
author: "Aine Fairbrother-Browne"
date: "03/22"
output: 
  html_document:
    theme: paper
    highlight: kate
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: false  
    fig_caption: true
---

```{r setup, include=FALSE}

# import libs
library(tidyverse)
library(DT)
library(ggforce)
library(scales)
library(lubridate)
library(ggrepel)

knitr::opts_chunk$set()
knitr::opts_knit$set(root.dir = "~/Documents/PhD/r_projects/ghost_flights/")

```

# Obtain data  

* I downloaded UK airline activity data from [here](https://www.caa.co.uk/data-and-analysis/uk-aviation-market/airlines/uk-airline-data/)  
* It is divided up by month, and each monthly dataset must be downloaded separately  
* For each month, I downloaded the following table: "Table 03 - All Services", using the command `wget -c https://www.caa.co.uk/Documents/Download/8634/87542ebf-3d02-44c1-8650-d8dbd6eca742/2863 -O dec_21_all_services.csv;`, for example    
* The inclusion criteria as per CAA for airlines and passengers contained in their data is as follows:  
* Airlines: "Airlines with an aircraft having a Maximum Take Off Mass (MTOM) above 40 tonnes and/or an A type Operating licence or Air Transport Licence are required, under Section 84 (1) of the Civil Aviation Act 1982, to report regular monthly data. However, charter services performed by aircraft below 15 tonnes MTOM and sole use/exempt in type are not included in published tables."  
* Passengers: "Recorded as 'uplifted passengers', they cover passengers carried by the airlines worldwide, not only to/from the UK, and relate to passengers who have paid the equivalent of 25% of the normal fare. Note that infants under 2 years are not included."  

```{bash eval=F, include=F}

cd /Users/ainefairbrother/Documents/PhD/r_projects/ghost_flights/data

# 2019 pre pandemic, non christmas - 80%
wget -c https://www.caa.co.uk/Documents/Download/4032/36ea4cce-bc59-4e62-9aad-32bfcc0e0a9e/1809 -O nov_2019_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4023/fe48c60b-e232-4390-b9de-a51ab6d29c8f/1716 -O dec_2019_all_services.csv
# 2020
wget -c https://www.caa.co.uk/Documents/Download/4161/064d174e-57fe-45b0-8e4b-7cfc56c608bf/1916 -O jan_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4159/d65c2023-d89c-4e79-be68-c09b27642dc4/1903 -O feb_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4152/0a3528d3-1299-4936-abd2-804e650af367/1957 -O mar_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4143/38f65663-dc9e-4c1a-b66e-df2e54e9151c/1862 -O apr_2020_all_services.csv
# summer 2020 - 0%
wget -c https://www.caa.co.uk/Documents/Download/4141/4b593b14-5e4b-4ebd-b0be-4d7006ea8d58/1968 -O may_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4132/5da78166-10f1-4f3a-ae56-408d8e2f5ff1/1944 -O jun_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4128/26d80e8a-1719-4c7c-9a24-4b16a848c0b4/1929 -O jul_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4127/62f55b7b-32a8-4f63-8004-b46f3cad3927/2891 -O aug_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4126/b43e6670-33a9-44f5-a604-032b361a55df/2010 -O sep_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4125/ec86e1dd-f1d2-4ac0-951d-b08dcc5aca90/1996 -O oct_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4124/83c23520-07f9-43af-a99c-63ada4f73d17/1983 -O nov_2020_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4091/34c32772-3a2a-4447-84e9-d4eef041ec2e/1890 -O dec_2020_all_services.csv
# 2021
wget -c https://www.caa.co.uk/Documents/Download/4194/04952113-6bc8-4d59-a2d1-c22138dd3a49/2049 -O jan_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4192/0435c01a-3064-4ce4-9e1d-86949f5d0286/2036 -O feb_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4191/f60e8ed0-ae6f-49a5-8e36-9df5d84fa314/2090 -O mar_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4190/31572161-42a1-4d5f-9c8a-2dc100f2227d/2023 -O apr_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4189/bc809c54-0164-4863-bd13-9c9acfeda975/2103 -O may_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/4169/52e37af5-22d5-4489-94f3-af11cc7e3f5b/2076 -O jun_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/3924/087de34d-6d50-45fb-933c-836af966a0b3/2062 -O jul_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/8389/766c7b54-84b9-40bc-b212-281cab5268d3/2688 -O aug_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/8390/3de520e0-b11c-4dae-a2b1-9f9c7c90735f/2728 -O sep_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/8391/90979cf6-8574-470a-8859-60727cf31602/2808 -O oct_2021_all_services.csv
# winter 2022 - 50% 
wget -c https://www.caa.co.uk/Documents/Download/8392/3105042e-778d-4833-b990-d794213b6bc6/2779 -O nov_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/8634/87542ebf-3d02-44c1-8650-d8dbd6eca742/2863 -O dec_2021_all_services.csv
wget -c https://www.caa.co.uk/Documents/Download/9085/957ac730-2947-4be0-b029-3191e6fb2a3e/2935 -O jan_2022_all_services.csv

```

* Read in downloaded data  

```{r message=FALSE, warning=FALSE, include=FALSE}

setwd("/Users/ainefairbrother/Documents/PhD/r_projects/ghost_flights/data/")
caa.dat = plyr::ldply(list.files(path="/Users/ainefairbrother/Documents/PhD/r_projects/ghost_flights/data/", pattern="*.csv"), read_csv)

```

* Do some `lubridate` magic to give `reporting_month`, `reporting_year` and `reporting_month_year` columns - this will make plotting easier  

```{r}

caa.dat = caa.dat %>% 
  dplyr::mutate(reporting_month=month(ym(reporting_period), label=T), 
                reporting_year=year(ym(reporting_period)),
                reporting_month_year=paste0(reporting_month,"-",reporting_year),
                reporting_period=ym(reporting_period)
  ) %>% 
  dplyr::relocate(rundate, reporting_period, reporting_month, reporting_year, reporting_month_year) %>% 
  dplyr::mutate(airline_name=gsub(" LTD", "", airline_name))

```

# Filter data  

* We will filter for only passenger services, as this is the focus of the project  

```{r}

caa.dat = caa.dat %>% 
  dplyr::filter(type_of_operations=="Passenger")

```

# Calculating variables  

I calculated the following variables:  
* Seat kilometers used as percent of available  
* Seat kilometers unused as percent of available  
* Average passengers per flight  

```{r}

caa.dat = caa.dat %>% 
  dplyr::mutate(
    
    # get raw values for those that are divided by 1000
    seat_km_available = seat_km_available_x1000*1000,
    seat_km_used = seat_km_used_x1000*1000,
    aircraft_km = aircraft_km_x1000*1000,
    
    # calculate used as a % of avail, and % unused
    seat_km_used_as_percent_of_avail = (seat_km_used/seat_km_available)*100,
    seat_km_unused_as_percent_of_avail = 100-seat_km_used_as_percent_of_avail,
    
    # calculate average passengers per flight
    avg_passengers_per_flight=total_passengers_uplifted/no_flights,
    
    # # calculate numbers of actual seats empty and full
    # actual_empty_seats = round((seat_km_used/aircraft_km), 0), 
    # actual_full_seats = round((seat_km_available/aircraft_km), 0), 
    # percent_full_seat_of_avail = (actual_empty_seats/(actual_empty_seats+actual_full_seats))*100
    
  ) %>% 
  
  # remove redundant cols
  dplyr::select(-contains("x1000"))

```

# Data example  

```{r}

caa.dat %>% 
  dplyr::select(-contains("cargo")) %>% 
  dplyr::select(-c(rundate, reporting_month, reporting_year, reporting_period, type_of_operations, seat_km_used_as_percent_of_avail, seat_km_unused_as_percent_of_avail, avg_passengers_per_flight)) %>% 
  dplyr::relocate(airline_name) %>% 
  write.csv("~/Documents/PhD/r_projects/ghost_flights/example_dat.csv")

```


# Plot data    
## Define slot rule variables  

* When were the slot rules in place? use it or loose it - https://www.routesonline.com/news/29/breaking-news/297416/uk-increases-minimum-slot-usage-threshold-for-summer-2022/  
* IATA Summer schedule - begins on the last Sunday of March and ends on the last Saturday of October.  
* IATA Winter schedule - begins on the last Sunday of October and ends on the last Saturday of March.  
* pre pandemic = 80%  
* However, in response to the pandemic, the European Commission waived the “use it or lose it” rule for the summer 2020 and winter 2020/21 seasons.  
* Following the end of the Brexit transition period—during which time the UK was following Brussels’ rules on slots—the UK chose to extend the waiver to cover the summer 2021 season and then introduced a 50% threshold for the winter 2021/22 season in decisions that mirrored those taken by the EC.  

* Nov 2019 - March 2020 = 80%  
* April 2020 - Oct 2020 = 0%  
* Nov 2020 - March 2021 = 0%  
* April 2021 - Oct 2021 = 0%  
* Nov 2021 - March 2022 = 50%  

* Define slot rule time boundaries  

```{r}

apr.20 = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2020-04-01"] %>% as.numeric(.)
nov.20 = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2020-11-01"] %>% as.numeric(.)
nov.21 = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2021-11-01"] %>% as.numeric(.)

```

## Define lockdown time variables    

* Define lockdown time boundaries  
* Data from https://www.instituteforgovernment.org.uk/charts/uk-government-coronavirus-lockdowns  

```{r}

# first lockdown 23/03 - August (2020)
lockdown.2020.1.start = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2020-04-01"] %>% as.numeric(.)
lockdown.2020.1.end = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2020-08-01"] %>% as.numeric(.)

# 2nd lockdown: 05/11/2020 - 02/12/2020  
lockdown.2020.2.start = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2020-11-01"] %>% as.numeric(.)
lockdown.2020.2.end = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2020-12-01"] %>% as.numeric(.)

# 3rd lockdown: 06/01/21 - may (2021)
lockdown.2021.3.start = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2021-01-01"] %>% as.numeric(.)
lockdown.2021.3.end = caa.dat %>% dplyr::pull(reporting_period) %>% unique() %>% .[.=="2021-05-01"] %>% as.numeric(.)

```

## Define data to plot - filter airlines  

* Focusing on the largest 5 airlines by passenger capacity
* Calculate airlines  

```{r}

caa.dat %>% 
  dplyr::pull(airline_name) %>% 
  unique() %>% 
  length() # 32 airlines 

```

```{r}

top.5.big.airlines = caa.dat %>% 
  # filter for airlines with data in every month
  dplyr::group_by(airline_name) %>%
  # allow 20% missing data, filter out airlines with more than this
  dplyr::filter(n() > as.numeric(caa.dat$reporting_month_year %>% unique() %>% length())-5 ) %>%  
  # to get the top 5 airlines by avg seat_km_available
  dplyr::summarise(mean_seat_km_available = mean(seat_km_available)) %>% 
  dplyr::arrange(-mean_seat_km_available) %>% 
  dplyr::pull(airline_name) %>% unique() %>% .[1:5] 

data.to.plot = caa.dat %>% 
  dplyr::filter(airline_name %in% top.5.big.airlines)

```

## Define colour palette  

```{r}

library(RColorBrewer)
# Define the number of colors you want
nb.cols = data.to.plot %>%  
  dplyr::pull(airline_name) %>% 
  unique() %>% 
  length() %>%
  as.numeric()

mycolors = colorRampPalette(brewer.pal(nb.cols, "Set1"))(nb.cols)

```

## Plot timecourse, no shading    

```{r, fig.width=8, fig.height=4}

p0 = data.to.plot %>% 
  ggplot(data=., aes(x=reporting_period, y=seat_km_available, colour=airline_name)) +
  theme_bw(base_size=11) +
  geom_line(alpha=0.5) +
  geom_point(aes(size=seat_km_used_as_percent_of_avail, alpha=seat_km_used_as_percent_of_avail)) +
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=10),
        legend.title = element_text(size=10),
        legend.direction = "horizontal", 
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5),
        axis.text.x = element_text(angle=55, hjust=1)) +
  labs(colour="Airline", 
       size="% passenger capacity used", 
       alpha="",
       title="") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  scale_colour_manual(values = mycolors) +
  scale_alpha(guide="none") +
  scale_x_date(breaks=scales::date_breaks("1 month")) +
  ylab("\nPassenger capacity available\n") +
  xlab("\nReporting period start date")

p0

```

```{r}

p0.pre.pandemic = data.to.plot %>% 
  
  dplyr::filter(reporting_month_year %in% c("Nov-2019", "Dec-2019", "Jan-2020")) %>% 
  
  ggplot(data=., aes(x=reporting_period, y=seat_km_available, colour=airline_name)) +
  theme_bw(base_size=11) +
  geom_line(alpha=0.5) +
  geom_point(aes(size=seat_km_used_as_percent_of_avail, alpha=seat_km_used_as_percent_of_avail)) +
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=10),
        legend.title = element_text(size=10),
        legend.direction = "horizontal", 
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5),
        axis.text.x = element_text(angle=55, hjust=1)) +
  labs(colour="Airline", 
       size="% passenger capacity used", 
       alpha="",
       title="") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  scale_colour_manual(values = mycolors) +
  scale_alpha(guide="none") +
  scale_x_date(breaks=scales::date_breaks("1 month")) +
  ylab("\nPassenger capacity available\n") +
  xlab("\nReporting period start date")

p0.pre.pandemic


```


## Plot capacity time-course with 0% slot rule highlighted  

```{r, fig.width=8, fig.height=4}

p1 = data.to.plot %>% 
  ggplot(data=., aes(x=reporting_period, y=seat_km_available, colour=airline_name)) +
  # shade 0% capacity rule area
  annotate("rect", xmin = as.Date("2020-04-01"), xmax = as.Date("2021-11-01"), ymin = -Inf, ymax = Inf, alpha = 0.1) +
  geom_vline(xintercept=apr.20, linetype=3, alpha=0.4) +
  geom_vline(xintercept=nov.21, linetype=3, alpha=0.4) +
  theme_bw(base_size=11) +
  geom_line(alpha=0.5) +
  geom_point(aes(size=seat_km_used_as_percent_of_avail, alpha=seat_km_used_as_percent_of_avail)) +
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=10),
        legend.title = element_text(size=10),
        legend.direction = "horizontal", 
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5),
        axis.text.x = element_text(angle=55, hjust=1)) +
  labs(colour="Airline", 
       size="% passenger capacity used", 
       alpha="",
       title="") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  scale_colour_manual(values = mycolors) +
  scale_alpha(guide="none") +
  scale_x_date(breaks=scales::date_breaks("1 month")) +
  ylab("\nPassenger capacity available\n") +
  xlab("\nReporting period start date") 

p1

```

## Plot capacity time-course with lockdowns highlighted  

```{r, fig.width=8, fig.height=4}

p2 = data.to.plot %>% 
  ggplot(data=., aes(x=reporting_period, y=seat_km_available, colour=airline_name)) +
  theme_bw(base_size=11) +
  # shade lockdown spans
  annotate("rect", xmin = as.Date("2020-04-01"), xmax = as.Date("2020-08-01"), ymin = -Inf, ymax = Inf, alpha = 0.1) +
  annotate("rect", xmin = as.Date("2020-11-01"), xmax = as.Date("2020-12-01"), ymin = -Inf, ymax = Inf, alpha = 0.1) +
  annotate("rect", xmin = as.Date("2021-01-01"), xmax = as.Date("2021-05-01"), ymin = -Inf, ymax = Inf, alpha = 0.1) +
  # add UK lockdown markers
  geom_vline(xintercept = lockdown.2020.1.start, linetype=3, alpha=0.4) +
  geom_vline(xintercept = lockdown.2020.1.end,   linetype=3, alpha=0.4) +
  geom_vline(xintercept = lockdown.2020.2.start, linetype=3, alpha=0.4) +
  geom_vline(xintercept = lockdown.2020.2.end,   linetype=3, alpha=0.4) +
  geom_vline(xintercept = lockdown.2021.3.start, linetype=3, alpha=0.4) +
  geom_vline(xintercept = lockdown.2021.3.end,   linetype=3, alpha=0.4) +
  # add points and lines
  geom_line(alpha=0.5) +
  geom_point(aes(size=seat_km_used_as_percent_of_avail, alpha=seat_km_used_as_percent_of_avail)) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=10),
        legend.title = element_text(size=10),
        legend.direction = "horizontal", 
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5),
        axis.text.x = element_text(angle=55, hjust=1)) +
  labs(colour="Airline", 
       size="% passenger capacity used", 
       alpha="",
       title="") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  scale_colour_manual(values = mycolors) +
  scale_alpha(guide="none") +
  scale_x_date(breaks=scales::date_breaks("1 month")) +
  ylab("\nPassenger capacity available\n") +
  xlab("\nReporting period start date")

p2

```

## Focus on between-lockdown space, plot used capacity vs. available capacity between Aug 2020 and Dec 2020   

```{r, fig.width=8, fig.height=4}

p3 = data.to.plot %>% 
  dplyr::filter(reporting_month_year %in% c("Aug-2020", "Sep-2020", "Oct-2020", "Nov-2020")) %>% 
  dplyr::distinct() %>% 
  ggplot(data=., aes(x=seat_km_used, y=seat_km_available, colour=airline_name, label=round(seat_km_used_as_percent_of_avail))) +
  theme_bw(base_size=10) +
  geom_abline(intercept = 0) +
  xlim(0, 4.2e+09) +
  ylim(0, 4.2e+09) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=9.5),
        legend.title = element_text(size=9.5),
        legend.direction = "horizontal",
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5), 
        axis.text.x = element_text(angle=90, hjust=0.5)) +
  labs(colour="Airline", size="% passenger capacity used") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  xlab("Capacity used") +
  ylab("Capacity available") +
  scale_colour_manual(values = mycolors) +
  coord_fixed(ratio = 1) +
  ggrepel::geom_label_repel() +
  facet_wrap(~factor(reporting_month_year, levels=c("Aug-2020", "Sep-2020", "Oct-2020", "Nov-2020")), nrow=2, ncol=3) 

p3

```

## Plot pre-pandemic (Nov 2019 - Jan 2022)  

```{r, fig.width=8, fig.height=4}

p6 = data.to.plot %>% 
  dplyr::filter(reporting_month_year %in% c("Nov-2019", "Dec-2019", "Jan-2020")) %>% 
  ggplot(data=., aes(x=seat_km_used, y=seat_km_available, colour=airline_name, label=round(seat_km_used_as_percent_of_avail))) +
  theme_bw(base_size=10) +
  geom_abline(intercept = 0) +
  xlim(0, 1.5e+10) +
  ylim(0, 1.5e+10) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=9.5),
        legend.title = element_text(size=9.5),
        legend.direction = "horizontal",
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5), 
        axis.text.x = element_text(angle=90, hjust=0.5)) +
  labs(colour="Airline", size="% passenger capacity used") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  xlab("Capacity used") +
  ylab("Capacity available") +
  scale_colour_manual(values = mycolors) +
  coord_fixed(ratio = 1) +
  facet_grid(~factor(reporting_month_year, levels=c("Nov-2019", "Dec-2019", "Jan-2020"))) +
  ggrepel::geom_label_repel()

p6

```

```{r}

data.to.plot %>% 
  # dplyr::filter(airline_name=="BRITISH AIRWAYS PLC") %>% 
  dplyr::filter(reporting_month_year %in% c("Nov-2019"))

```

## Plot 1st lockdown

```{r, fig.width=8, fig.height=4}

p6.1 = data.to.plot %>% 
  dplyr::filter(reporting_month_year %in% c("Apr-2020", "May-2020", "Jun-2020", "Jul-2020", "Aug-2020")) %>% 
  ggplot(data=., aes(x=seat_km_used, y=seat_km_available, colour=airline_name, label=round(seat_km_used_as_percent_of_avail))) +
  theme_bw(base_size=10) +
  geom_abline(intercept = 0) +
  xlim(0, 4.2e+09) +
  ylim(0, 4.2e+09) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=9.5),
        legend.title = element_text(size=9.5),
        legend.direction = "horizontal",
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5), 
        axis.text.x = element_text(angle=90, hjust=0.5)) +
  labs(colour="Airline", size="% passenger capacity used") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  xlab("Capacity used") +
  ylab("Capacity available") +
  scale_colour_manual(values = mycolors) +
  coord_fixed(ratio = 1) +
  facet_wrap(~factor(reporting_month_year, levels=c("Apr-2020", "May-2020", "Jun-2020", "Jul-2020", "Aug-2020")), nrow=2) +
  ggrepel::geom_label_repel()

p6.1

```

## Plot 3rd lockdown

```{r, fig.width=8, fig.height=4}

p6.2 = data.to.plot %>% 
  dplyr::filter(reporting_month_year %in% c("Jan-2021", "Feb-2021", "Mar-2021", "Apr-2021", "May-2021")) %>% 
  dplyr::distinct() %>% 
  ggplot(data=., aes(x=seat_km_used, y=seat_km_available, colour=airline_name, label=round(seat_km_used_as_percent_of_avail))) +
  theme_bw(base_size=10) +
  geom_abline(intercept = 0) +
  xlim(0, 3.5e+09) +
  ylim(0, 3.5e+09) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=9.5),
        legend.title = element_text(size=9.5),
        legend.direction = "horizontal",
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5), 
        axis.text.x = element_text(angle=90, hjust=0.5)) +
  labs(colour="Airline", size="% passenger capacity used") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  xlab("Capacity used") +
  ylab("Capacity available") +
  scale_colour_manual(values = mycolors) +
  coord_fixed(ratio = 1) +
  ggrepel::geom_label_repel() +
  facet_wrap(~factor(reporting_month_year, levels=c("Jan-2021", "Feb-2021", "Mar-2021", "Apr-2021", "May-2021")), nrow=2) 


p6.2

```

## Plot post-pandemic (Jul 2021 - Jan 2022)  

```{r, fig.width=8, fig.height=4}

p4 = data.to.plot %>% 
  dplyr::filter(reporting_month_year %in% c("Oct-2021", "Nov-2021", "Dec-2021", "Jan-2022")) %>% 
  dplyr::distinct() %>% 
  ggplot(data=., aes(x=seat_km_used, y=seat_km_available, colour=airline_name, label=round(seat_km_used_as_percent_of_avail))) +
  theme_bw(base_size=10) +
  geom_abline(intercept = 0) +
  xlim(0, 8.9e+09) +
  ylim(0, 8.9e+09) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=9.5),
        legend.title = element_text(size=9.5),
        legend.direction = "horizontal",
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5), 
        axis.text.x = element_text(angle=90, hjust=0.5)) +
  labs(colour="Airline", size="% passenger capacity used") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  xlab("Capacity used") +
  ylab("Capacity available") +
  scale_colour_manual(values = mycolors) +
  coord_fixed(ratio = 1) +
  ggrepel::geom_label_repel() +
  facet_wrap(~factor(reporting_month_year, levels=c("Oct-2021", "Nov-2021", "Dec-2021", "Jan-2022")), nrow=3)

p4

```

```{r}

p4.1 = data.to.plot %>% 
  dplyr::filter(reporting_month_year %in% c("Oct-2021", "Nov-2021", "Dec-2021", "Jan-2022")) %>% 
  dplyr::distinct() %>% 
  ggplot(data=., aes(x=reporting_period, y=seat_km_available, colour=airline_name, label=round(seat_km_used_as_percent_of_avail))) +
  theme_bw(base_size=11) +
  # add points and lines
  geom_line(alpha=0.5) +
  geom_point(aes(size=seat_km_used_as_percent_of_avail, alpha=seat_km_used_as_percent_of_avail)) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=10),
        legend.title = element_text(size=10),
        legend.direction = "horizontal", 
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5),
        axis.text.x = element_text(angle=55, hjust=1)) +
  labs(colour="Airline", 
       size="% passenger capacity used", 
       alpha="",
       title="") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  scale_colour_manual(values = mycolors) +
  scale_alpha(guide="none") +
  scale_x_date(breaks=scales::date_breaks("1 month")) +
  ylab("\nPassenger capacity available\n") +
  xlab("\nReporting period start date") +
  ggrepel::geom_label_repel() 

p4.1

```


```{r, fig.width=8, fig.height=4}

p5 = data.to.plot %>% 
  dplyr::filter(airline_name!="BRITISH AIRWAYS PLC") %>% 
  dplyr::filter(reporting_month_year %in% c("Jul-2021", "Aug-2021", "Sep-2021", "Oct-2021", "Nov-2021", "Dec-2021", "Jan-2022")) %>% 
  ggplot(data=., aes(x=seat_km_used, y=seat_km_available, colour=airline_name)) +
  theme_bw(base_size=9.5) +
  geom_abline(intercept = 0) +
  geom_point(aes(size=seat_km_used_as_percent_of_avail)) +
  xlim(0, 3.5e+09) +
  ylim(0, 3.5e+09) +
  # set theme
  theme(legend.position =  "top",
        legend.background = element_blank(),
        legend.box.background = element_blank(),
        legend.text = element_text(size=9.5),
        legend.title = element_text(size=9.5),
        legend.direction = "horizontal",
        legend.box = "vertical",
        plot.title = element_text(hjus=0.5), 
        axis.text.x = element_text(angle=90, hjust=0.5)) +
  labs(colour="Airline", size="% passenger capacity used") +
  guides(colour = guide_legend(order = 1), size=guide_legend(order = 3)) +
  xlab("\nCapacity used\n") +
  ylab("\nCapacity available\n") +
  scale_colour_manual(values = mycolors) +
  coord_fixed(ratio = 1) +
  facet_grid(~factor(reporting_month_year, levels=c("Jul-2021", "Aug-2021", "Sep-2021", "Oct-2021", "Nov-2021", "Dec-2021", "Jan-2022")))

p5

```

# Other findings

## Understanding how many empty planes the unused capacity is  

```{r, fig.width=3, fig.height=1.5}

p7 = data.to.plot %>% 
  dplyr::filter(reporting_month_year %in% c("Nov-2019", "Dec-2019", "Jan-2020")) %>% 
  dplyr::mutate(est.empty.planes = (seat_km_unused_as_percent_of_avail/100)*no_flights) %>% 
  dplyr::select(airline_name, reporting_month_year, reporting_period, no_flights, est.empty.planes) %>% 
  dplyr::arrange(-est.empty.planes) %>% 
  
  ggplot(data=., aes(x=factor(reporting_month_year, levels=c("Nov-2019", "Dec-2019", "Jan-2020")), y=est.empty.planes, fill=airline_name)) +
  theme_bw() +
  geom_col(position="dodge") +
  scale_fill_manual(values = mycolors) +
  labs(fill="Airline") +
  xlab("") +
  ylab("\nNumber of empty planes\n")
  
p7

```





## Highlighting flights running with < 10 passengers per flights  

```{r}

caa.dat %>% 
  dplyr::filter( (seat_km_used_as_percent_of_avail<=15)) %>% 
  dplyr::group_by(reporting_year, airline_name) %>% 
  dplyr::summarise(sum_flights_less_10_avg_passengers = sum(no_flights)) %>% 
  dplyr::arrange(-sum_flights_less_10_avg_passengers)


```

# Save plots  

```{r}

setwd("~/Documents/PhD/r_projects/ghost_flights/img/")

ggsave(
  filename="p1.png",
  plot = p1,
  device = "png",
  scale = 2,
  width = 5,
  height = 3,
  units = c("in"),
  dpi = 500
)
ggsave(
  filename="p2.png",
  plot = p2,
  device = "png",
  scale = 2,
  width = 5,
  height = 3,
  units = c("in"),
  dpi = 500
)
# inter-lockdown space, 2 rows, 4 facets
ggsave(
  filename="p3.png",
  plot = p3,
  device = "png",
  scale = 2,
  width = 5,
  height = 4,
  units = c("in"),
  dpi = 500
)
ggsave(
  filename="p4.png",
  plot = p4,
  device = "png",
  scale = 2,
  width = 5,
  height = 5,
  units = c("in"),
  dpi = 500
)
ggsave(
  filename="p4_1.png",
  plot = p4.1,
  device = "png",
  scale = 2,
  width = 5,
  height = 3,
  units = c("in"),
  dpi = 500
)
ggsave(
  filename="p5.png",
  plot = p5,
  device = "png",
  scale = 2,
  width = 5,
  height = 5,
  units = c("in"),
  dpi = 500
)
ggsave(
  filename="p6.png",
  plot = p6,
  device = "png",
  scale = 2,
  width = 5.5,
  height = 3,
  units = c("in"),
  dpi = 500
)
# First lockdown plot - 2 rows, 5 facets
ggsave(
  filename="p6_1.png",
  plot = p6.1,
  device = "png",
  scale = 2,
  width = 5,
  height = 4,
  units = c("in"),
  dpi = 500
)
# Third lockdown plot - 2 rows, 5 facets
ggsave(
  filename="p6_2.png",
  plot = p6.2,
  device = "png",
  scale = 2,
  width = 5,
  height = 4,
  units = c("in"),
  dpi = 500
)
ggsave(
  filename="p7.png",
  plot = p7,
  device = "png",
  scale = 2,
  width = 5,
  height = 2.5,
  units = c("in"),
  dpi = 500
)

```

# Notes

## Explaining the data/ measures  

* Seat km available is a measure of the number of seats and the distances those seats could fly as determined by scheduled flights and aircraft capacity  
* So if an airline has 1 Boeing 737-800 (capacity 189) and has 1 scheduled flight from London to Edinburgh (approx. 500km), then it will have an available seat km of 189*500=94500  
* This makes seat km available a good inter-airline comparison metric, as it takes into account the variable fleets of airlines, as well as the variable haulages of different airlines  
* Each month, the CAA releases its UK Airline dataset, which includes monthly totals for passengers uplifted, distance flown, seat kilometers available, seat kilometers used etc., for each airline  

## Timecourse plot explainer  

* This plot shows a timecourse spanning from November 2019 (pre-pandemic) to January 2022 (latest figures released by the CAA), with dates plotted along the x-axis corresponding to the monthly data-release schedule of the CAA  
* The y-axis gives the % of seat kilometers used of those available to the airline  
* The shaded area represents the duration of the UK waiver on the minimum slot rule, i.e. there was no minimum activity required by an airline during this time to retain their airport slot  
* The points are set to be large and opaque when % seat km used is high, and small and transparent when % seat km is low  
* So, it is clear to see when availability meets or outstrips demand  
* The availability is shown on the y-axis, and a dark opaque point will show that a high % of this was used, conversely, a small transparent point will show that a low % of this was used  
  
# Session info
```{r}

sessionInfo()

```



